Resultants and subresultants of $p$-adic polynomials

### 2015-07-21 resultants/resultants.tex
The stabilization process in general

@@

@@

\subsubsection*{The notion of lattice}

@@

Although the defintion of a lattice is similar to that of $\Z$-lattice 
in $\R^d$, the geometrical representation of it is quite different since 
the elements of $\A$ themselves are not equidistributed in $K$ but form 
a ball inside $K$ (they are exactly elements of norm $\leq 1$). More 
generally, assume that $E$ is equipped with a ultrametric norm $\Vert 
\cdot \Vert_E$ compatible with that on $K$ (\emph{i.e.} $\Vert \lambda x 
\Vert_E = |\lambda| \cdot \Vert x \Vert_E$ for $\lambda \in K$, $x \in 
E$). (A typical example is $E = K^n$ equipped with the sup norm.) One
checks that the subsets
$$B_E(r) = \big\{ \,\, x \in E \quad|\quad \Vert x \Vert_E \leq r \,\,\big\}$$
are all lattices in $E$. Moreover, any lattice is deduced from $B_E(1)$ 
by applying a bijective linear endomorphism of $E$. Therefore, lattices 
should be thought as special neighborhoods of $0$. As a consequence, 
cosets of the form $x + H$, where $H$ is a lattice, appear as 
interesting candidates to model precision. This feeling is consolidated 
by the following result which roughly speaking claims that such cosets 
behave quite nicely under differentiable maps.

@@

\subsubsection*{Application to precision}

\todo{Add a subsection to explain the stabilization process in
general.}

>>>>

\definecolor{beige}{rgb}{0.9,0.9,0.7}

@@

\renewcommand{\min}{\text{\rm min}}
\renewcommand{\max}{\text{\rm max}}

@@

\subsubsection{The notion of lattice}

@@

Although the defintion of a lattice is similar to that of $\Z$-lattice 
in $\R^d$, the geometrical representation of it is quite different since 
the elements of $\A$ themselves are not equidistributed in $K$ but form 
a ball inside $K$ (they are exactly elements of norm $\leq 1$). More 
generally, assume that $E$ is equipped with a ultrametric norm $\Vert 
\cdot \Vert_E$ compatible with that on $K$ (\emph{i.e.} $\Vert \lambda x 
\Vert_E = |\lambda| \cdot \Vert x \Vert_E$ for $\lambda \in K$, $x \in 
E$). (A typical example is $E = K^n$ equipped with the sup norm.) One
checks that the subsets
$$B_E(r) = \big\{ \,\, x \in E \quad|\quad \Vert x \Vert_E \leq r \,\,\big\}$$
are all lattices in $E$. Moreover, any lattice is deduced from $B_E(1)$ 
by applying a bijective linear endomorphism of $E$. Therefore, lattices 
should be thought as special neighborhoods of $0$ (see Figure~\ref{fig:lattice}).
%
\begin{figure}
\hfill
\begin{tikzpicture}
\begin{scope}[beige, rounded corners=10pt]
\fill (0.1,0.2) rectangle (3.9,5.8);
\end{scope}

\begin{scope}[xshift=2cm,yshift=3cm,rotate=20]
\draw[black,fill=green!50] (-0.7,-1.5) rectangle (0.7,1.5);
\draw[black,->] (-1,0)--(1,0);
\draw[black,->] (0,-2)--(0,2);
\end{scope}

\begin{scope}[very thick, rounded corners=10pt]
\draw (0.1,0.2) rectangle (3.9,5.8);
\end{scope} 
\end{tikzpicture}
\hfill \null

\caption{Picture of a lattice in the ultrametric world}
\label{fig:lattice}
\end{figure}
%
As a consequence, cosets of the form $x + H$, where $H$ is a lattice, 
appear as interesting candidates to model precision. This feeling is 
consolidated by the following result which roughly speaking claims that 
such cosets behave quite nicely under differentiable maps.

@@

\subsubsection{Application to precision}

Let us now briefly explain how one can use Lemma~\ref{lem:padicprec} for 
tracking precision.

\paragraph{Tracking precision \emph{locally}}

Assume first that we want to perform a given rather simple operation --- 
corresponding, say, to an elementary step (\emph{e.g.} an iteration of 
the main loop) of the algorithm we are executing --- modeled by a 
stricty differentiable function $g$ defined on an open subset $U$ of a 
normed $K$-vector space $E$ and taking values in another normed 
$K$-vector space $F$.
Our input is an approximated element of $U$ which is represented 
as a coset $C$ with respect to some lattice $H$, that is a subset of $U$ 
of the form $C = x+H$ for some $x \in U$. We would like to insist on the 
following: the value of $x$ is a priori \emph{not} given; only is given 
the subset $C$. However, since $H$ is stable under addtion, we have $C = 
x+H$ \emph{any} element $x \in C$.\footnote{This assertion means that 
any element of the ``rectangle'' $C$ is a center of it... which might be 
surprising if we are accustomed to real numbers.} As explained in \S 
\ref{subsubsec:computDVR}, assuming that $g$ is given as an algebraic 
expression, the naive solution for evaluating $g(C)$ consists in using 
formulas \eqref{eq:precadd}--\eqref{eq:precdiv}. However, this often 
results in an overestimation on the precision, in the following sense: 
this method leads to some inclusion
$$g(C) = g(x+H) \subset y + H_{\text{naive}}$$
where $y \in F$ and $H_{\text{naive}}$ is a lattice which is generally 
much more larger that the best one, namely $g'(x)(H)$ according to
Lemma~\ref{lem:padicprec} (assuming that the assumptions of this Lemma
are fullfiled). In order to avoid this and be sharp on
precision, another solution consists in splitting the computation of 
$g(C)$ into two parts as follows:
\begin{enumerate}[(A)]
\item \label{item:gpx} compute $g'(x)(H)$, and
\item \label{item:gx} compute $g(x)$ for some $x \in C$.
\end{enumerate}
Part~\eqref{item:gpx} is not easy to handle in full generality: in order 
to be efficient, a special close analysis taking advantage of the 
particular problem under consideration is often necessary. For the 
example considered in the paper (which is that of subresultants), we will 
use a \emph{theoretical} argument in order to exhibit a good approximation 
of $g'(x)(H)$, that are lattices $H_\min$ and $H_\max$ such that:
\begin{equation}
\label{eq:Hminmax}
H_\min \subset g'(x)(H) \subset H_\max
\end{equation}
for a negligible cost. Up to allowing in some cases a \emph{very slight} 
leak of precision, this would be enough for the application we have in 
mind. We refer to \S \ref{subsec:stabilization} for more details about
this part.

For now on, we assume that $H_\min$ and $H_\max$ satisfying 
Eq.~\eqref{eq:Hminmax} have been computed and we focus on 
part~\eqref{item:gx}, which also requires some discussion. Indeed, 
computing $g(x)$ is not straightforward because $x$ itself lies in a 
$K$-vector space and therefore cannot be stored and manipulated on a 
computer. Nevertheless, one can take advantage of the fact that $x$ may 
be chosen arbitratily in $C$. More precisely, we pick a sublattice $H'$ 
of $H$ and consider the new approximated element $x+H' \subset x+H$. 
Concretely, this means that we arbitrarily increase the precision on the 
given input $x$. Now, applying the naive method with $x+H'$, we compute 
some $y \in F$ and some lattice $H'_{\text{naive}} \subset F$ with the 
property that:
$$g(x+H') \subset y + H'_{\text{naive}}.$$
If furthemore $H'$ is chosen in such a way that $H'_{\text{naive}} 
\subset H_\min$, the two cosets $y + g'(x)(H)$ and $g(C)$ have a 
non-empty intersection because $g(x)$ lies in both. Therefore they 
must coincide. We deduce that $y \in g(C)$. This exactly means that
$y$ is an acceptable value for $g(x)$ and we are done.
Moreover, estimating the dependance of $H'_{\text{naive}}$ in terms 
of $H'$ is usually rather easy (remember that $f$ is supposed to model
a simple operation). Hence since $H_\min$ is known --- as
we had assumed --- finding $H'$ satisfying the required assumption is
generally not difficult. 
The whole method is illustrating on Figure~\ref{fig:adaptative}.

\begin{figure}
\hfill
\begin{tikzpicture}[scale=1.2]

\begin{scope}[beige, rounded corners=10pt]
\fill (0.1,0.2) rectangle (3.9,5.8);
\fill (6.1,0.2) rectangle (9.9,5.8);
\end{scope}

\draw[->,very thick] (4.1,3)--(5.9,3)
  node[midway,above] { $g$ };

\begin{scope}[xshift=2cm,yshift=3.5cm,rotate=20]
\draw[black,fill=green!50] (-0.8,-1.8) rectangle (0.8,1.8);
\node[scale=0.75] at (0,0) { $x{+}H$ };
\end{scope}
\draw[black,fill=purple!50] (2,2.8) rectangle (2.8,2);
\node[scale=0.75] at (2.4,2.4) { $x{+}H'$ };

\begin{scope}[xshift=8cm,yshift=3cm]
\draw[black,fill=orange!50] (-1.7,-2.3) rectangle (1.7,1.8);
\node[above left,scale=0.75] at (1.7,-2.3) { $g(x){+}H_\max$ };
\draw[black,fill=green!50,rotate=-40] (-1,-1.3) rectangle (1,1.3);
\node[scale=0.75] at (0.15,1.2) { $g(x){+}{}$ };
\node[scale=0.75] at (0.1,0.9) { $g'(x)(H)$ };
\draw[black,fill=red!50] (-0.8,-0.8) rectangle (0.4,0.4);
\node[scale=0.75] at (-0.2,0.1) { $g(x){+}H_\min$ };
\draw[black,fill=purple!50] (-0.7,-0.7) rectangle (0.3,-0.3);
\node[scale=0.75] at (-0.2,-0.5) { $y{+}H'_{\text{naive}}$ };
\end{scope}

\begin{scope}[very thick, rounded corners=10pt]
\draw (0.1,0.2) rectangle (3.9,5.8);
\draw (6.1,0.2) rectangle (9.9,5.8);
\end{scope} 
\end{tikzpicture}
\hfill \null

\caption{Method for tracking precision based on Lemma~\ref{lem:padicprec}}
\label{fig:adaptative}
\end{figure}

\paragraph{Tracking precision \emph{globally}}

As already said, we shall use the above method for tracking precision 
while executing a single step in a complete algorithm. Let us now 
address the problem of ``glueing''. We consider an algorithm \texttt{F} 
consisting in a succession of $n$ steps $\texttt{G}_0, \ldots,
\texttt{G}_{n-1}$. It is modeled by a strictly differentiable function $f : 
U \to F$ where $U$ is an open subset in a normed $K$-vector space $E$ 
and $F$ is a normed $K$-vector space. The input of \texttt{F} is an 
approximated element in $U$ represented as a coset $C = x+H$ where $x 
\in U$ and $H$ is a lattice.
We also introduce notations for each individual step. For all $i$, we
assume that $\texttt{G}_i$ is modeled by a strictly differentiable
function $g_i : U_i \to U_{i+1}$ where $U_i$ is an open subset is 
some normed $K$-vector space $E_i$ and, by convention, $U_0 = U$, $E_0 
= E$ and $U_n = E_n = F$. We thus have:

$$f = g_{n-1} \circ g_{n-2} \circ \cdots \circ g_1 \circ g_0.$$
For all $i$, we set $f_i = g_{i-1} \circ \cdots \circ g_0$. It is the 
function modeling the execution of the $i$ first steps of our algorithm. 
We further define $x_i = f_i(x)$ and $H_i = f_i'(x)(H)$. The chain rule 
for composing differentials readily implies the recurrence
\begin{equation}
\label{eq:giprime}
H_{i+1} = g_i'(x_i)(H_i)
\end{equation}
For simplicity, we make the following assumptions:
\begin{itemize}
\item the $\Zp$-submodule $H_i$ is a lattice in $E_i$ such that $x_i + 
H_i \subset U_i$;
\item the triple $(g_i, x_i, H_i)$ satisfies the assumptions of 
Lemma~\ref{lem:padicprec};
\item for all $i$, we have succeeded in finding (good enough) explicit 
lattices $H_{\min,i}$ and $H_{\max,i}$ such that $H_{\min,i} \subset H_i 
\subset H_{\max,i}$;
\item for all $i$, we have succeeded in finding an explicit lattice
$H'_i$ such that, while tracking naively precision, we end up with an
inclusion
$$g_i(x_i + H'_i) = x_{i+1} + H_{\text{naive},i+1}$$
with $H_{\text{naive},i+1} \subset H_{\min,i+1}$.
\end{itemize}
We note that the first and the second assumptions are quite strong 
because it implies in particular that the sequence of $\dim E_i$ is 
non-increasing. However, it really simplifies the forthcoming discussion 
and will be harmless for the application developed in this paper. As
already mentionned, the construction of $H_{\min,i}$ and $H_{\max,i}$
will generally follow from a \emph{theoretical} argument depending on
the setting, while exhibiting $H'_i$ will often be straightforward.
Anyway, we are now in position to apply the method for 
tracking precision locally we have discussed earlier to all $g_i$'s. 
This leads to a \emph{stabilized} version of the algorithm \texttt{G} 
whose skeleton is depicted in Algorithm~\ref{algo:stabilizedG}.
%
\begin{algorithm}
  \SetKwInOut{Input}{Input}
  \SetKwInOut{Output}{Output}
  \Input{$x$ given at precision $O(H)$}
  \Output{$g(x)$ given at precision $O(H_{\max,n})$}

  $x_0 \leftarrow x$;

  \For{$i = 0, \ldots, n-1$}
    {\textbf{lift} $x_i$ to precision $O(H'_i)$;

     $x_{i+1} \leftarrow \texttt{F}_i(x_i)$;}

  \Return{$x_n + O(H_{\max,n})$};
\caption{Stabilized version of \texttt{G}}
\label{algo:stabilizedG}
\end{algorithm}
%
